Approximating the end of nested sampling

### 2023-07-28 paper/paper.tex
Fixed figures...

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously accomplishes the two tasks of Bayesian inference, model comparison and parameter estimation \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{split}. The core algorithm is unique for being the only algorithm which estimates high-dimensional volumes using \textit{order statistics}, which makes high-dimensional integration feasible. It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
\par
Many instances of software for nested sampling exist, including popular implementations such as \textsc{MultiNest} \citep{multinest} and \textsc{PolyChord} \citep{polychord} which provide flexible options for nested sampling runs, as well as post-processing packages such as \textsc{anesthetic} \citep{anesthetic} to obtain and plot sample statistics. However, currently no such software is able to estimate how long a given run should last. This is of high importance to the end user, who has no indication in the middle of a run whether it will take minutes, or weeks and months to finish; even an order of magnitude estimate is useful compared to the status quo of complete ignorance.
\par
This project sets out a novel approach for estimating the endpoint of a nested sampling run at an intermediate stage. The idea is to predict the form of the likelihood function in the region we have yet to sample from, using the information we have already gained from our previous samples. We begin in section \ref{sec:Background} with an overview of Bayesian inference and nested sampling. Section \ref{sec:Endpoint prediction} presents the theory and methodology of making endpoint predictions, before results for both toy and real examples are shown in section \ref{sec:Results}. Finally, conclusions are made in section \ref{sec:Conclusions}.

@@

\subsection{Uncertainties and the posterior bulk}\label{subsec:Uncertainties and the posterior bulk}
\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/LX.pdf}
\end{center}
\caption{Plot showing the likelihood and the posterior weights $\Like(X)X$. The bulk of posterior mass is contained within a region of size $e^{-\DKL}$, meaning one must iterate through at least $n\DKL$ steps to fully explore the posterior.}
\label{fig:LX}
\end{figure}

\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/X_samples.pdf}
\end{center}
\caption{Plot showing samples of $\log X$ plotted against the likelihoods of the sample points. The uncertainty increases with the number of steps taken.}
\label{fig:X_samples}
\end{figure}

@@

\subsection{Likelihood approximation method}
We want to make an estimate of the number of iterations until the termination condition is met, which is when the remaining evidence falls below a fraction $\epsilon$ of the accumulated evidence.
Predictions can be made ahead of time because of the live points, which are sampled from higher likelihood regions that we have not yet compressed to. It is important to note that there will always be a small region which the live points do not penetrate because they are too diffuse, which can be seen in figure \ref{fig:dead_live_points}. 
\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/dead_live_points.pdf}
\end{center}
\caption{Plots showing the state of a nested sampling at an intermediate iteration. The live points are samples from higher likelihood regions, which allow us to see the remaining likelihood ahead of time. However, there is always a small high likelihood region where we have no samples, simply because the live points are too diffuse.}
\label{fig:dead_live_points}
\end{figure}
\par
If we could estimate the form of $\Like(X)$ at the higher likelihoods in terms of some \textit{likelihood model} $f(X, \thetab)$, we would be able to express the prior volume $X_f$ which we need to compress to as
\begin{equation}
	\Delta Z = \epsilon Z_\mathrm{tot},
\end{equation}
\begin{equation}\label{model endpoint}
	\int_0^{X_f} f(X, \thetab)\ dX = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ dX + Z_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $Z_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_f$ can then be identified by solving the above equation either analytically or numerically. Remembering that at each iteration $\log X$ decreases by $1/\nlive$, the total number of iterations $N_f$ will be
\begin{equation}
	N_f = - \nlive \log X_f .
\end{equation}

@@

\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/gauss_overfit.pdf}
\end{center}
\caption{Plot showing fits of the likelihood corresponding to different draws of $\log X$. The different realisations of the likelihood lead to a set of endpoint predictions, from which a mean and variance can be obtained.}
\label{fig:gauss_overfit}
\end{figure}

\section{Results}\label{sec:Results}
\subsection{Toy data}
Let us now apply the methodology developed above in practice. We consider the following toy examples:
\begin{enumerate}[leftmargin=0.3in, label=(\alph*)]
	\item A spherically symmetric Gaussian, $\log \Like = -|\bm{r}|^2/2\sigma^2$ in $d$ dimensions
	\item A discontinous "wedding cake" likelihood with Gaussian profile and parameter $\alpha_\mathrm{w}$ controlling the plateau depth, as presented in \cite{plateaus} and shown in figure \ref{fig:wedding_3d}.
	\item A Cauchy likelihood $\log \Like = - \frac{1 + d}{2} \log (1 + |\bm{r}|^2/\gamma^2)$.
\end{enumerate}
\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/wedding_3d.pdf}
\end{center}
\caption{An example ``wedding cake" likelihood which has plateaus on top of a Gaussian profile}
\label{fig:wedding_3d}
\end{figure}

Parameters were chosen to have large values of $\DKL$, which are more difficult to model due to the large distance to the posterior bulk.
\par
Nested sampling runs were made for the above likelihoods for $\nlive = 500$ (plots for different numbers of live points are not shown here, but the results are broadly the same). The state at each iteration (i.e. the likelihoods of the dead and live points) was then extracted using \textsc{anesthetic}. At each iteration, the model is fitted over a number of points determined by the bandwidth selection procedure outlined above. For the error bars, we draw an \textit{ensemble} of 25 $\bm{X}$ samples from the distribution \eqref{xi_formula}, and find the optimal $\thetab$ and corresponding  $N_f$ for each sample to get the mean and variance of the prediction. 
\par
Results are shown in figure \ref{fig:toy_predictions}. For all cases, the predictions become more accurate as the iteration increases because the samples shrink around and eventually cross the posterior bulk, leading to an increasingly representative picture of the posterior. The method works very well for the first two likelihoods, recovering the correct endpoint within standard error for the entire run. This is expected, since the model likelihood has the same form as the true likelihoods, so one expects the extrapolation made to be accurate even when we are far away from the posterior bulk.
\par
The Cauchy likelihood shows the result when this is not the case; without samples in the bulk, a Gaussian extrapolation performs poorly. However, the correct order of magnitude is still obtained, which is a significant improvement on the status quo of complete ignorance. 
\begin{figure*}
\begin{center}
	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics[scale=]{Figures/gauss_predictions.pdf}}
	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics[scale=]{Figures/wedding_predictions.pdf}}
	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics[scale=]{Figures/cauchy_predictions.pdf}}
\end{center}
\caption{Plot showing endpoint predictions for the toy likelihoods. The mean estimate is dark blue, while the shading shows the $1\sigma$ and $2\sigma$ uncertainties. For the first two likelihoods (which have Gaussian profile), the true endpoint falls within error for practically all of the run. The Cauchy likelihood is less well-estimated because it is highly non-Gaussian, with the true endpoint outside the reasonable uncertainty range but in the correct order of magnitude.}
\label{fig:toy_predictions}
\end{figure*}

\subsection{Real data}
The model was tested on real nested sampling runs for quantification of cosmological parameter tensions. Results are shown in figure \ref{fig:lcdm_predictions}, with the likelihood approximation method compared against the benchmark.
\par
The likelihood model clearly performs better. For all of the chains, the true endpoint was recovered within arund standard error by the halfway point, and falls within the correct order of magnitude throughout all of the runs. In comparison, the benchmark model mostly failed to get the correct order of magnitude on the number of iterations remaining, especially near the end of each run.
\par
Furthermore, the first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian likelihood model. For all chains, further work needs to be done to place a more accurate upper bound on the prediction uncertainty at early iterations. 
\begin{figure*}
\begin{center}
	\includegraphics[scale=]{Figures/lcdm_predictions.pdf}
\end{center}
\caption{Plots showing the endpoint predictions for the benchmark model as well as the likelihood approximation model. The likelihood model visibly performs better, managing to recover the true endpoint within around standard error at the halfway point for all the chains. The first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian model.}
\label{fig:lcdm_predictions}
\end{figure*}

>>>>

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously accomplishes the two tasks of Bayesian inference, model comparison and parameter estimation \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{split}. The core algorithm is unique for being the only algorithm which estimates volumes using \textit{order statistics}, which makes high-dimensional integration feasible. It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
\par
Many instances of software for nested sampling exist, including popular implementations such as \textsc{MultiNest} \citep{multinest} and \textsc{PolyChord} \citep{polychord} which provide flexible options for nested sampling runs, as well as post-processing packages such as \textsc{anesthetic} \citep{anesthetic} to obtain and plot sample statistics. However, currently no such software is able to estimate how long a given run should last. This is of high importance to the end user, who has no indication in the middle of a run whether it will take minutes, or weeks and months to finish; even an order of magnitude estimate is useful compared to the status quo of complete ignorance.
\par
This project sets out a novel approach for estimating the endpoint of a nested sampling run at an intermediate stage. The idea is to predict the form of the likelihood function in the region we have yet to sample from, using the information we have already gained from our previous samples. We begin in section \ref{sec:Background} with an overview of Bayesian inference and nested sampling. Section \ref{sec:Endpoint prediction} presents the theory and methodology of making endpoint predictions, before results for both toy and real examples are shown in section \ref{sec:Results}. Finally, conclusions are made in section \ref{sec:Conclusions}.

@@

\subsection{Uncertainties and the posterior bulk}\label{subsec:Uncertainties and the posterior bulk}
\begin{figure}
\begin{center}
	\includegraphics{Figures/LX.pdf}
\end{center}
\caption{Plot showing the likelihood and the posterior weights $\Like(X)X$. The bulk of posterior mass is contained within a region of size $e^{-\DKL}$, meaning one must iterate through at least $n\DKL$ steps to fully explore the posterior.}
\label{fig:LX}
\end{figure}

\begin{figure}
\begin{center}
	\includegraphics{Figures/X_samples.pdf}
\end{center}
\caption{Plot showing samples of $\log X$ plotted against the likelihoods of the sample points. The uncertainty increases with the number of steps taken.}
\label{fig:X_samples}
\end{figure}

@@

\subsection{Likelihood approximation method}
We want to make an estimate of the number of iterations until the termination condition is met, which is when the remaining evidence falls below a fraction $\epsilon$ of the accumulated evidence.
Predictions can be made ahead of time because of the live points, which are sampled from higher likelihood regions that we have not yet compressed to. It is important to note that there will always be a small region which the live points do not penetrate because they are too diffuse, which can be seen in figure \ref{fig:dead_live_points}. 
\begin{figure}
\begin{center}
	\includegraphics{Figures/dead_live_points.pdf}
\end{center}
\caption{Plots showing the state of a nested sampling at an intermediate iteration. The live points are samples from higher likelihood regions, which allow us to see the remaining likelihood ahead of time. However, there is always a small high likelihood region where we have no samples, simply because the live points are too diffuse.}
\label{fig:dead_live_points}
\end{figure}
\par
If we could estimate the form of $\Like(X)$ at the higher likelihoods in terms of some \textit{likelihood model} $f(X, \thetab)$, we would be able to express the prior volume $X_f$ which we need to compress to as
\begin{equation}
	\Delta Z = \epsilon Z_\mathrm{tot},
\end{equation}
\begin{equation}\label{model endpoint}
	\int_0^{X_f} f(X, \thetab)\ dX = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ dX + Z_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $Z_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_f$ can then be identified by solving the above equation either analytically or numerically. Remembering that at each iteration $\log X$ decreases by $1/\nlive$, the total number of iterations $N_f$ will be
\begin{equation}
	N_f = - \nlive \log X_f .
\end{equation}

@@

\begin{figure}
\begin{center}
	\includegraphics{Figures/gauss_overfit.pdf}
\end{center}
\caption{Plot showing fits of the likelihood corresponding to different draws of $\log X$. The different realisations of the likelihood lead to a set of endpoint predictions, from which a mean and variance can be obtained.}
\label{fig:gauss_overfit}
\end{figure}

\section{Results}\label{sec:Results}
\subsection{Toy data}
Let us now apply the methodology developed above in practice. We consider the following toy examples:
\begin{enumerate}[leftmargin=0.3in, label=(\alph*)]
	\item A spherically symmetric Gaussian, $\log \Like = -|\bm{r}|^2/2\sigma^2$ in $d$ dimensions
	\item A discontinous "wedding cake" likelihood with Gaussian profile and parameter $\alpha_\mathrm{w}$ controlling the plateau depth, as presented in \cite{plateaus} and shown in figure \ref{fig:wedding_3d}.
	\item A Cauchy likelihood $\log \Like = - \frac{1 + d}{2} \log (1 + |\bm{r}|^2/\gamma^2)$.
\end{enumerate}
\begin{figure}
\begin{center}
	\includegraphics{Figures/wedding_3d.pdf}
\end{center}
\caption{An example ``wedding cake" likelihood which has plateaus on top of a Gaussian profile}
\label{fig:wedding_3d}
\end{figure}

Parameters were chosen to have large values of $\DKL$, which are more difficult to model due to the large distance to the posterior bulk.
\par
Nested sampling runs were made for the above likelihoods for $\nlive = 500$ (plots for different numbers of live points are not shown here, but the results are broadly the same). The state at each iteration (i.e. the likelihoods of the dead and live points) was then extracted using \textsc{anesthetic}. At each iteration, the model is fitted over a number of points determined by the bandwidth selection procedure outlined above. For the error bars, we draw an \textit{ensemble} of 25 $\bm{X}$ samples from the distribution \eqref{xi_formula}, and find the optimal $\thetab$ and corresponding  $N_f$ for each sample to get the mean and variance of the prediction. 
\par
Results are shown in figure \ref{fig:toy_predictions}. For all cases, the predictions become more accurate as the iteration increases because the samples shrink around and eventually cross the posterior bulk, leading to an increasingly representative picture of the posterior. The method works very well for the first two likelihoods, recovering the correct endpoint within standard error for the entire run. This is expected, since the model likelihood has the same form as the true likelihoods, so one expects the extrapolation made to be accurate even when we are far away from the posterior bulk.
\par
The Cauchy likelihood shows the result when this is not the case; without samples in the bulk, a Gaussian extrapolation performs poorly. However, the correct order of magnitude is still obtained, which is a significant improvement on the status quo of complete ignorance. 
\begin{figure*}
\begin{center}
	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics{Figures/gauss_predictions.pdf}}
	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics{Figures/wedding_predictions.pdf}}
	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics{Figures/cauchy_predictions.pdf}}
\end{center}
\caption{Plot showing endpoint predictions for the toy likelihoods. The mean estimate is dark blue, while the shading shows the $1\sigma$ and $2\sigma$ uncertainties. For the first two likelihoods (which have Gaussian profile), the true endpoint falls within error for practically all of the run. The Cauchy likelihood is less well-estimated because it is highly non-Gaussian, with the true endpoint outside the reasonable uncertainty range but in the correct order of magnitude.}
\label{fig:toy_predictions}
\end{figure*}

\subsection{Real data}
The model was tested on real nested sampling runs for quantification of cosmological parameter tensions. Results are shown in figure \ref{fig:lcdm_predictions}, with the likelihood approximation method compared against the benchmark.
\par
The likelihood model clearly performs better. For all of the chains, the true endpoint was recovered within arund standard error by the halfway point, and falls within the correct order of magnitude throughout all of the runs. In comparison, the benchmark model mostly failed to get the correct order of magnitude on the number of iterations remaining, especially near the end of each run.
\par
Furthermore, the first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian likelihood model. For all chains, further work needs to be done to place a more accurate upper bound on the prediction uncertainty at early iterations. 
\begin{figure*}
\begin{center}
	\includegraphics{Figures/lcdm_predictions.pdf}
\end{center}
\caption{Plots showing the endpoint predictions for the benchmark model as well as the likelihood approximation model. The likelihood model visibly performs better, managing to recover the true endpoint within around standard error at the halfway point for all the chains. The first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian model.}
\label{fig:lcdm_predictions}
\end{figure*}
